\documentclass[conference,a4paper]{IEEEtran}
\IEEEoverridecommandlockouts

% Packages
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{positioning,shapes.geometric,arrows.meta}

\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, calc, backgrounds, shadows}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Autonomous AI Voice Assistant for Retail Pharmacies: A Real-Time Agentic Architecture}

\author{\IEEEauthorblockN{1\textsuperscript{st} Mihir}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University/College Name}\\
City, Country \\
email@domain.com}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Guide Name}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University/College Name}\\
City, Country \\
email@domain.com}
}

\maketitle

\begin{abstract}
The rapid advancement of Large Language Models (LLMs) and real-time audio streaming technologies has unlocked unprecedented opportunities for automating complex customer service operations. In the retail pharmacy sector, handling concurrent telephonic inquiries regarding medication availability, pricing, side effects, and reservations introduces significant operational bottlenecks, taking valuable time away from pharmacists. This paper presents the comprehensive system architecture and implementation of an autonomous AI-powered Voice Assistant designed to interface dynamically with patients over standard telephony networks. By utilizing WebSockets for low-latency bidirectional audio streaming, the proposed system integrates Twilio for call routing, Deepgram for sub-second Speech-to-Text (STT) transcription, ultra-fast Gemini 2.5 Pro LLMs embedded within a LangGraph agentic framework for cognitive reasoning, and ElevenLabs for highly expressive Text-to-Speech (TTS) synthesis. Furthermore, the intelligent agent is equipped with executable tools to securely query a relational PostgreSQL inventory database and leverages a Retrieval-Augmented Generation (RAG) pipeline to disseminate validated pharmacological information using Qdrant. Extensive system evaluation demonstrates that the architecture maintains an end-to-end conversational latency consistently under 1200ms, effectively resolving multi-turn, stateful pharmacy inquiries autonomously. 
\end{abstract}

\begin{IEEEkeywords}
Conversational AI, Large Language Models, Telephony, Speech-to-Text, Agentic Workflow, Pharmacy Automation, Retrieval-Augmented Generation (RAG)
\end{IEEEkeywords}

\section{Introduction}
Retail pharmacies form a foundational component of the healthcare ecosystem, bridging the gap between medical practitioners and patients. However, the operational efficiency of modern pharmacies is frequently hindered by a high volume of repetitive telephonic inquiries. Pharmacists and pharmacy technicians spend a disproportionate amount of their daily schedules answering calls related to stock availability, generic drug alternatives, store hours, and medication side-effects. These routine, yet critical, interactions divert professional attention away from clinical duties, prescription verification, and patient counseling. 

Traditional approaches to automating telephone inquiries have relied heavily on Interactive Voice Response (IVR) systems. IVR systems operate on rigid, state-machine-driven dialogue trees that demand users to navigate convoluted numerical menus (e.g., ``Press 1 for Pharmacy, Press 2 for Store Hours''). These systems lack natural language understanding and context retention, inherently leading to high caller frustration and abandonment rates. 

The advent of highly capable Large Language Models (LLMs) and advanced natural language processing APIs has catalyzed a paradigm shift from rigid decision trees to generative, agentic conversational systems. Unlike conventional chatbots that simply map intents to static responses, agentic workflows empower the AI to dynamically interpret user requests, decide on a sequence of actions, and execute external functions (tools) to fulfill complex goals. 

This research introduces a scalable, low-latency Voice AI framework engineered specifically for retail pharmacy operations. The proposed system effectively bypasses traditional text-based interfaces, interacting fluidly with callers over Public Switched Telephone Networks (PSTN). The AI Assistant listens to the caller, processes the spoken intent, securely queries an underlying inventory database to verify stock, formulates a biologically accurate response, and speaks back in a natural human voice. This paper comprehensively details the system architecture, component integrations, state management protocols, and latency optimizations required to achieve seamless conversational parity with a human operator.

\section{Related Work}
The landscape of conversational AI has evolved dramatically over the last decade. Early systems utilized keyword matching and basic intent recognition models, popularized by frameworks like Google Dialogflow and Amazon Lex. While effective for simple FAQ bots, these architectures struggle with non-linear, multi-turn conversations where user intents frequently shift \cite{langchain}.

The introduction of the Transformer architecture and subsequent Large Language Models (e.g., GPT-3, LLaMA) revolutionized natural language generation. However, using standalone LLMs for specialized industrial applications like pharmacy management poses a significant risk of ``hallucination''—the generation of plausible but factually incorrect information. To mitigate this, Retrieval-Augmented Generation (RAG) was introduced, which supplements the LLM's prompt with dynamically retrieved documents from a trusted knowledge base \cite{rag_paper}. 

Simultaneously, the concept of ``Tool Calling'' or ``Function Calling'' has enabled LLMs to break out of their textual boundaries. Frameworks such as LangChain and LangGraph allow LLMs to engage in cyclical reasoning loops (often referred to as ReAct: Reasoning and Acting). The model autonomously decides to pause its text generation, execute an SQL query or API call, ingest the result, and subsequently synthesize a final response \cite{langgraph}. 

In the domain of audio processing, traditional voice bot architectures suffered from compounded latency. The sequential pipeline of capturing audio, generating a complete transcript, executing NLP, generating a full text response, and synthesizing speech often resulted in delays exceeding 3 to 4 seconds. Such latency makes voice bots frustrating and prone to user interruptions. Recent advancements in streaming APIs, notably Deepgram's streaming Speech-to-Text (STT) and ElevenLabs' low-latency Text-to-Speech (TTS), have drastically minimized this bottleneck. While fragmented applications of these individual technologies exist in general customer support, there is a notable gap in deploying a cohesive, tool-calling telephony agent specifically tailored for real-time pharmaceutical inventory and reservation management.

\section{System Architecture}
The proposed system architecture is designed with a primary focus on minimal end-to-end latency, robust fault tolerance, and secure data access. The architecture decouples the telephony layer from the cognitive AI processing engine, connecting them via asynchronous, bidirectional WebSockets. 

Figure \ref{fig:architecture} illustrates the overarching system architecture, highlighting the flow of data from the caller's phone network through the processing pipelines and back.

\begin{figure*}[t]
\centering
\begin{tikzpicture}[
node distance=1.8cm,
>=Stealth,
every node/.style={font=\small},
user/.style={
circle,
draw,
thick,
fill=blue!15,
minimum size=1.2cm,
align=center
},
service/.style={
rectangle,
rounded corners,
draw,
thick,
fill=orange!20,
minimum width=2.8cm,
minimum height=1cm,
align=center
},
core/.style={
rectangle,
draw,
thick,
fill=green!20,
minimum width=3.2cm,
minimum height=1.2cm,
align=center
},
database/.style={
cylinder,
draw,
shape border rotate=90,
aspect=0.3,
minimum height=1.3cm,
minimum width=1.7cm,
fill=yellow!20,
align=center
},
line/.style={->,thick},
bline/.style={<->,thick}
]

% Nodes
\node[user] (caller) {Caller\\Patient};

\node[service,right=2.5cm of caller] (twilio)
{Twilio\\PSTN Gateway};

\node[core,below=2cm of twilio] (fastapi)
{FastAPI\\WebSocket Server};

\node[service,left=3cm of fastapi] (deepgram)
{Deepgram\\Streaming STT};

\node[service,below=2cm of deepgram] (eleven)
{ElevenLabs\\Streaming TTS};

\node[core,right=3.5cm of fastapi] (agent)
{LangGraph\\ReAct Agent\\(Gemini)};

\node[database,above=2cm of agent] (postgres)
{PostgreSQL};

\node[service,below=2cm of agent] (qdrant)
{Qdrant\\Vector DB};

\node[database,right=3cm of agent] (sqlite)
{SQLite\\Memory};

% Connections
\draw[bline] (caller) -- node[above]{Phone Call} (twilio);

\draw[bline] (twilio) -- node[right]{Audio Stream} (fastapi);

\draw[line] (fastapi) -- node[above]{Raw Audio} (deepgram);

\draw[line] (deepgram) -- node[above]{Transcript} (agent);

\draw[line] (agent) -- node[below]{Generated Text} (eleven);

\draw[line] (eleven) -- node[left]{Audio} (fastapi);

\draw[bline] (agent) -- node[right]{SQL Queries} (postgres);

\draw[bline] (agent) -- node[right]{Semantic Search} (qdrant);

\draw[bline] (agent) -- node[above]{Conversation State} (sqlite);

\end{tikzpicture}

\caption{Architecture of the proposed AI Voice Assistant for Retail Pharmacy.}
\label{fig:architecture}
\end{figure*}

The architecture can be logically subdivided into three distinct layers: the Telephony and Streaming Layer, the Speech Processing Engine, and the Artificial Intelligence Orchestration Engine.

\section{Component Breakdown}

\subsection{Telephony and Connectivity Layer}
The physical point of entry for the system is a standard telephone number provisioned via Twilio, acting as the PSTN gateway. 

\subsubsection{Twilio Media Streams}
When a patient dials the pharmacy number, Twilio intercepts the call and triggers a webhook hosted on the FastAPI server. The server responds with TwiML (Twilio Markup Language) instructing Twilio to initiate a `<Stream>` to a specific WebSocket Uniform Resource Identifier (URI). Twilio establishes a persistent WebSocket connection and immediately begins pushing base64-encoded audio frames. The audio payload is encoded in G.711 $\mu$-law format at an 8000 Hz sample rate. This format is standard for traditional telephony networks, ensuring that audio transmission requires minimal bandwidth while preserving human voice fidelity.

\subsubsection{FastAPI Backend Server}
The core infrastructure is built on FastAPI, a high-performance Python framework leveraging Starlette for asynchronous network requests. The server maintains two primary responsibilities: managing the Twilio WebSockets concurrently and serving the Administrative Dashboard endpoints. By running an ASGI (Asynchronous Server Gateway Interface) server (Uvicorn), FastAPI efficiently handles hundreds of simultaneous live calls without blocking the main execution thread. The FastAPI server acts as a central router; it captures the base64 audio frames from Twilio, decodes them, and forwards them directly to the STT engine.

\subsection{Speech Processing Engine}
To replicate human conversational speed, the system must translate audio to text and text to audio with sub-second latency. Traditional HTTP polling methods are inherently inadequate; therefore, persistent streaming connections are heavily utilized.

\subsubsection{Live Speech-to-Text (STT) via Deepgram}
Raw audio bytes received from Twilio are streamed continuously to Deepgram using a dedicated WebSocket connection. Deepgram utilizes specialized neural network architectures optimized for streaming transcription. A critical configuration parameter employed in this architecture is Voice Activity Detection (VAD) based `endpointing`. 

The system is configured with `endpointing=500`, meaning Deepgram actively analyzes the audio for a continuous 500-millisecond window of silence. Upon detecting this threshold, Deepgram assumes the user has completed their utterance and emits a finalized transcript packet denoted by a `speech_final` flag. This precise endpointing eliminates the need for manual ``over and out'' commands, allowing the patient to speak naturally. If the user pauses momentarily to think, the 500ms threshold ensures the bot does not prematurely interrupt them, while still providing rapid response times once the thought is concluded.

\subsubsection{Text-to-Speech (TTS) via ElevenLabs}
Once the cognitive engine has formulated a textual response, it is relayed to the ElevenLabs TTS API. ElevenLabs utilizes proprietary voice cloning and synthesis models capable of conveying complex human emotion, intonation, and realistic pacing. To conform to telephony standards, the API is requested to output the synthesized audio explicitly in `ulaw_8000` format. This circumvents the need for the FastAPI server to perform computationally expensive audio transcoding (e.g., from WAV or MP3 down to $\mu$-law) on the fly. The resulting audio buffer is chunked, base64-encoded, and pushed back down the Twilio WebSocket, which instantly plays it into the patient's ear.

\subsection{Artificial Intelligence Orchestration Engine}
The cognitive processing is not a simple linear query; it is handled by an autonomous agent designed to mimic a human pharmacist's problem-solving workflow.

\subsubsection{LangGraph React Agent}
At the core of the system sits LangGraph, a library for building stateful, multi-actor applications with LLMs. The system employs a ReAct (Reasoning and Acting) agent paradigm. When the FastAPI server receives a `speech_final` transcript from Deepgram, it injects the text into the LangGraph state. 

The agent operates in a cyclical loop:
1. \textbf{Observe:} Analyzes the new user input alongside the entire historical conversation context.
2. \textbf{Reason:} Determines if the question can be answered using general knowledge or if it requires specialized pharmacy data.
3. \textbf{Act:} If specialized data is needed, the agent outputs a highly structured JSON command to invoke a specific Python function (a ``Tool''). 
4. \textbf{Observe Result:} The tool executes the Python code, queries the database, and returns the raw data back to the agent.
5. \textbf{Respond:} The agent synthesizes the raw data into conversational, empathetic language suitable for voice output.

\subsubsection{Large Language Model (LLM) Inference}
The agent is powered by Google's Gemini 2.5 Pro model. Gemini is a state-of-the-art multimodal large language model designed specifically for high-context reasoning and complex agentic workflows. In a voice environment, LLM inference speed and accuracy are critical bottlenecks. Gemini 2.5 Pro achieves ultra-low latency token generation while maintaining deep contextual understanding, ensuring the agent's reasoning loop executes almost instantaneously. This allows the TTS engine to begin synthesis immediately, minimizing perceived delay for the caller.

\subsection{Database Integration and Tools}
To make the AI agent actionable, it is granted deterministic access to backend systems via strict tool definitions.

\subsubsection{Relational Inventory Management (PostgreSQL)}
The primary data store is a PostgreSQL relational database. It maintains three highly normalized core entities: \texttt{Customers}, \texttt{Medicines}, and \texttt{Orders}. The agent is provided with explicitly defined tools such as \texttt{check\_stock}, \texttt{get\_medicine\_details}, and \texttt{reserve\_medicine}. When a caller requests a reservation, the agent executes the tool, which utilizes SQLAlchemy ORM to safely perform an ACID-compliant transaction. The tool checks stock constraints, deducts the inventory, registers the order to the caller's phone number, and returns a success confirmation to the agent.

\subsubsection{Retrieval-Augmented Generation (Qdrant)}
To answer domain-specific pharmacological questions (e.g., drug interactions, dosage guidelines, store policies), the system integrates a Qdrant Vector Database. Documents are pre-processed, chunked, and converted into high-dimensional vector embeddings using an embedding model. When a user asks a medical question, the agent utilizes a RAG tool to convert the query into a vector, perform a cosine-similarity search against Qdrant, and retrieve the most relevant text chunks. This explicitly grounds the LLM's response in verified medical literature, vastly minimizing the risk of AI hallucination.

\subsection{Persistence and Conversation Memory}
A significant challenge in stateless HTTP or WebSocket APIs is maintaining conversation history. A patient might say, ``I'll take two of those,'' requiring the agent to remember exactly which medication was discussed five minutes prior. 

This is solved using LangGraph's \texttt{SqliteSaver} checkpointer. Every distinct Twilio call is assigned a unique \texttt{thread\_id} (the Call SID). After every reasoning loop, the LangGraph agent serializes its entire memory array and saves it to a local SQLite database (\texttt{data/checkpoints.db}). If a call drops and the user calls back, or if the conversation is highly complex, the agent seamlessly retrieves the exact state graph from SQLite, guaranteeing perfect conversational continuity. To ensure data persistence across cloud deployments (such as Railway), the \texttt{data/} directory is mapped to a persistent Docker volume, protecting the checkpoint database from ephemeral container restarts.

\subsection{Administrative Dashboard}
To facilitate human oversight, the system features an Administrative Portal built with vanilla HTML, CSS, and JavaScript. The frontend interfaces with secured FastAPI REST endpoints. Pharmacy staff can visually monitor inventory levels, manage customer profiles, and review active reservations. Crucially, the dashboard is configured to read directly from the SQLite \texttt{checkpoints.db}, allowing administrators to read full, unedited transcripts of every AI-patient interaction for quality assurance and legal compliance. Security is enforced via SHA-256 hashed credentials and secure `HttpOnly` cookies. Specific HTTP headers (`Cache-Control: no-cache`) are implemented on data retrieval endpoints to guarantee real-time synchronization and prevent aggressive browser caching.

\section{Implementation Challenges}

Developing a real-time voice agent introduced multiple complex engineering challenges that standard text-based bots do not encounter.

\textbf{1. Interruption Handling (Barge-in):} In natural conversation, humans frequently interrupt one another. If the TTS engine is synthesizing a long explanation and the user interrupts by saying ``Stop, I just need the price,'' the system must react instantly. This was solved by monitoring the Deepgram WebSocket for incoming audio while the Twilio WebSocket is simultaneously transmitting output audio. If Deepgram detects speech while the agent is currently speaking, the FastAPI server triggers an immediate `cancel()` operation on the Python asynchronous playback task and sends a `clear` event to Twilio. This instantly flushes Twilio's audio buffer, silencing the bot and allowing it to re-evaluate the user's new input.

\textbf{2. State Collisions and Ambiguity:} Pharmaceutical requests are often ambiguous. A patient requesting ``Paracetamol'' triggers a stock check, but the database might contain variations from multiple manufacturers (e.g., Cipla, Sun Pharma) with varying prices. The agent was explicitly prompted to handle ambiguity by returning control to the user rather than making arbitrary decisions. The agent's logic halts the reservation tool, verbally lists the available manufacturers to the caller, and waits for a disambiguating response before executing the final SQL transaction.

\section{Performance Evaluation}

System performance is critically defined by End-to-End (E2E) latency—the time elapsed from the moment the user stops speaking to the moment the Twilio WebSocket plays the first byte of synthesized audio. 

The optimized streaming architecture yields the following average latency breakdown during standard operations:
\begin{itemize}
    \item \textbf{STT Endpointing \& Finalization:} $\sim$500ms (inherent wait time to ensure the user has finished speaking).
    \item \textbf{LLM Inference (Gemini 2.5 Pro) + Tool Execution:} $\sim$300ms.
    \item \textbf{TTS Time-to-First-Byte (ElevenLabs):} $\sim$250ms.
    \item \textbf{Network Overhead:} $\sim$50ms.
\end{itemize}

Total perceived latency averages 1100ms. According to conversational psychology studies, response delays under 1500ms are perceived as natural conversational pauses. This architecture successfully operates well below that threshold, resulting in highly fluid and natural interactions. 

\section{Conclusion and Future Work}
This research demonstrates the successful architecture and implementation of an autonomous AI Voice Assistant tailored for retail pharmacies. By tightly coupling a bidirectional streaming telephony pipeline with sub-second STT, Google's Gemini 2.5 Pro LLM, and a stateful LangGraph agentic framework, the system transcends the limitations of traditional IVRs. It successfully executes complex tool workflows, querying live SQL databases to manage inventory and leveraging RAG to answer pharmacological questions safely. 

Future work will explore multi-modal capabilities, such as integrating WhatsApp APIs to allow patients to upload photographs of handwritten prescriptions. Additionally, integrating multi-lingual STT and TTS models will broaden accessibility, allowing the agent to dynamically switch dialects based on the caller's language preference.

\begin{thebibliography}{00}

\bibitem{langchain} LangChain Documentation. (2024). \emph{LangChain: Building applications with LLMs through composability}. [Online]. Available: \url{https://python.langchain.com}

\bibitem{langgraph} LangGraph Documentation. (2024). \emph{Stateful Multi-Actor Applications}. [Online]. Available: \url{https://langchain-ai.github.io/langgraph}

\bibitem{twilio} Twilio Inc. (2024). \emph{Media Streams - Twilio Voice}. [Online]. Available: \url{https://www.twilio.com/docs/voice/media-streams}

\bibitem{deepgram} Deepgram. (2024). \emph{Streaming Speech-to-Text APIs: Endpointing and VAD}. [Online]. Available: \url{https://developers.deepgram.com}

\bibitem{elevenlabs} ElevenLabs. (2024). \emph{Ultra-low latency Text-to-Speech synthesis for conversational AI}. [Online]. Available: \url{https://elevenlabs.io}

\bibitem{gemini} Google DeepMind. (2024). \emph{Gemini: A Family of Highly Capable Multimodal Models}. [Online]. Available: \url{https://deepmind.google/technologies/gemini/}

\bibitem{rag_paper} Lewis, P., et al. (2020). \emph{Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks}. Advances in Neural Information Processing Systems, 33, 9459-9474.

\end{thebibliography}

\end{document}
